### 2022-12-07 alignpair_letter.tex
Letter format draft

% title goes here:
\begin{flushleft}
% {\Large\textbf{COATi: statistical pairwise aligner robust to artifacts}}
{\Large\textbf{COATi: statistical pairwise alignment of protein coding sequences}}
% TODO: change name?
% codon-aware pairwise aligner + artifacts
\newline
% authors go here:
\\
Juan J. Garcia Mesa\textsuperscript{1,2},
Reed A. Cartwright\textsuperscript{1,3}
\\
\bigskip
1 Biodesign Institute, Arizona State University
\\
2 Ira A. Schools of Engineering, Arizona State University
\\
3 School of Life Sciences, Arizona State University
\\
\bigskip
* corresponding@author.mail

@@

% \abstract{Abstract}
\begin{abstract}
\noindent \textbf{Summary:} COATi is a statistical codon-aware pairwise aligner
that supports complex insertion-deletion models and is able to handle artifacts
present in genomic data.\\  % abstract summary can be longer.
\textbf{Availability:} The source code for COATi, along with documentation, is
freely available on GitHub: \url{https://github.com/CartwrightLab/coati} and is
implemented in C++.\\
\textbf{Supplementary information:} %\green{TODO}.
\end{abstract}

@@

\input{description}

\input{methods}

\input{results}

\section{Discussion}

COATi is an FST-based application that can calculate the optimal alignment
between a pair of sequences in the presence of artifacts using a statistical
model.
It will allow researchers to analyze more data with higher accuracy and
facilitate the study of important biological processes that shape genomic data.

>>>>

% title goes here:
\begin{flushleft}
{\Large\textbf{COATi: statistical pairwise alignment of protein coding sequences}}
\newline
% authors go here:
\\
Juan J. Garcia Mesa\textsuperscript{1,2},
Ziqi Zhu\textsuperscript{1,3},
Reed A. Cartwright\textsuperscript{1,3}
\\
\bigskip
1 Biodesign Institute, Arizona State University
\\
2 Ira A. Schools of Engineering, Arizona State University
\\
3 School of Life Sciences, Arizona State University
\\
\bigskip
* corresponding@author.mail

@@

\input{abstract}

@@

\input{materials_methods}

\input{results_discussion}

\section*{Availability}
The source code for COATi, along with documentation, is
freely available on GitHub: \url{https://github.com/CartwrightLab/coati} and is
implemented in C++.
Code to replicate the analysis can be found on GitHug: \url{https://github.com/jgarciamesa/TODO}.

% \section*{Supplementary information}
% \green{TODO}.

### 2022-12-07 figures/fig-evolution-fst.tex
Add marginal, update results
Letter format draft

\draw[line] (start1) -- (S1);
\draw[line] (S1) -- node[w] {$g$} (I1);
\draw[line] (S1) -- node[w] {$1-g$} (W1);
\draw[line] (U1) to[out=150,in=30]  node[w,above] {$e$} (I1);
\draw[line] (U1) -- node[w,pos=0.1,right] {$1-e$} (W1);
\draw[line] (W1) -- node[w] {$g$} (D1);
\draw[line] (W1) -- node[w] {$1-g$} (M1);
\draw[line] (V1) to[out=150,in=30] node[w,above] {$e$} (D1);
\draw[line] (D1) to[out=60,in=90] (end1 |- D1.south) to (end1);
\draw[line] (V1) -- node[w,swap] {$1-e$} (M1);
\draw[line] (M1) --  (end1);

@@

\draw[line] (M1) to[bend left=45] node[e,pos=0.25] {YYY\textrightarrow{}ZZZ} (S1);

>>>>

\draw[line] (start1) -- (S1);
\draw[line] (S1) -- node[w] {$g$} (I1);
\draw[line] (S1) -- node[w] {$1-g$} (W1);
\draw[line] (U1) to[out=150,in=30]  node[w,above] {$e$} (I1);
\draw[line] (U1) -- node[w,pos=0.1,right] {$1-e$} (W1);
\draw[line] (W1) -- node[w] {$g$} (D1);
\draw[line] (W1) -- node[w] {$1-g$} (M1);
\draw[line] (M1) -- (end1);
\draw[line] (V1) to[out=150,in=30] node[w,above] {$e$} (D1);
\draw[line] (D1) to[out=60,in=90] node[w] {$1-e$} (end1 |- D1.south) to (end1);
\draw[line] (V1) -- node[w,swap] {$1-e$} (M1);

@@

\draw[line] (M1) to[bend left=45] node[e,pos=0.25] {Y\textrightarrow{}Z} (S1);

### 2022-12-07 introduction.tex
Letter format draft

\section{Introduction}

% sequence alignment is important and heavily used
Sequence alignment is a cornerstone step in bioinformatics
\parencite{sequence_alignment_rosenberg_2009}.
% errors in genomic data that lead to erroneous downstream analyses
Uncorrected errors in sequence alignment can lead to erroneous results in
functional and comparative genomic studies \parencite{estimates_schneider_2009}.
% to correct this common practice is to discard lots of data
Given that errors and artifacts are common in molecular data, this requires
costly curation practices that discard large amounts of information.
% despite available aligners, there is room for improvement
% summarize sentence into ~ "alignment is performed in the amino acid space"
In addition, a common strategy is to perform alignment inference in the amino
acid space \parencite{bininda2005transalign,abascal2010translatorx}.
While this approach is an improvement over DNA models, it discards information,
underperforms compared to alignment at the codon level, and fails in the
presence of artifacts such as frameshifts and early stop codons.
Although some aligners incorporate codon substitution models, they do not
support frameshifts or lack as statistical model.

% to address this, we present COATi
To address this problem, we present COATi, short for COdon-Aware Alignment
Transducer, a statistical pairwise aligner that incorporates codon substitution
models and is robust to artifacts present in genomic data.

>>>>

\section*{Introduction}

Sequence alignment is a fundamental task in bioinformatics and a cornerstone step in comparative and functional genomic analysis \parencite{sequence_alignment_rosenberg_2009}.
Modern sequence analysis began with the heuristic homology algorithms of Needleman and Wunsch in 1970 \parencite{identification_smith_1981} and, while the methods developed since then have improved, alignment inference is not a solved problem \parencite{art_morrison_2015}.

A common strategy is to perform alignment inference in the amino acid space \parencite{bininda2005transalign,abascal2010translatorx}.
While this approach is an improvement over DNA models, it discards information, underperforms compared to alignment at the codon level, and fails in the presence of artifacts such as frameshifts and early stop codons.
Although some aligners incorporate codon substitution models, they do not support frameshifts or lack a statistical model.
In addition, while modeling indels to appear within codons is rare, this is often the case (Zhu \& Cartwright, personal communication, 2019).
Considering gaps to only appear between codons can result in missing the optimal alignment and inflate estimates of sequence divergence (Fig. \ref{fig:aln}).

\begin{figure}[h!]
% \begin{framed}
    \begin{minipage}[c]{0.65\textwidth}
        \includegraphics[scale=1]{fig-aln.pdf}
    \end{minipage}\hfill
    \begin{minipage}[c]{0.35\textwidth}
        \caption{
        \textbf{Standard algorithms produce suboptimal alignments.}
        (a) shows a possible alignment of an ancestor (A) sequence and a descendant (D) sequence.
        (b), (c), and (d) are the results of different aligners.
        Nucleotide mismatches are highlighted in red, phase zero, one, and two indels are shown in gray, purple, and orange, respectively.
        Phase one indel (purple) is also of type I (synonymous) while phase two indel (orange) is of type II (nonsynonymous). 
        COATi is able to retrieve the biological alignment (highlighted in blue).
        }
    \label{fig:aln}
    \end{minipage}
% \end{framed}
\end{figure}

Uncorrected errors in the alignment stage can lead to erroneous results in comparative and functional genomic studies \parencite{estimates_schneider_2009}.
Current methods are ill-equipped to handle common artifacts in genomic data, requiring costly curation practices that discard significant amounts of information.
To address this problem, we present COATi, short for COdon-aware Alignment Transducer, a pairwise statistical aligner that incorporates codon substitution models and is robust to artifacts present in modern genomic data.
